---
title: "DAM TSG Project assignment - Analysis of relation between salinity and temperature"
author: "Karenina Schröder"
date: "2025-08-29"

output: 
  html_document:
    number_sections: TRUE
    toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Relation between temperature and salinity in TSG data from DAM Underway project

```{r}
# read file tsg_data_cleaned into dataframe
data <- read.delim("tsg_data_cleaned.txt")
```

```{r}
sal <- data$Sal
temp <- data$Temp...C.
```

## Plotting salinity and temperature

### Simple plot of salinity vs. temperature

```{r}
plot(sal, temp)
```
--> seems not to show linear relation

### simple plot using ggplot 

```{r}
library(ggplot2)
ggplot(mapping= aes(sal, temp)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

## Fitting a linear model of salinity and temperature values

```{r}
mod <- lm(temp~sal)
summary(mod)
```

# Interpretation of results


**summary(mod) returns:**


![](./output_summary(mod).png)


## Clarifying summary(mod) output


- Intercept (3.77): When sal = 0, the predicted temp is 3.77.

- sal coefficient (0.474): For every 1-unit increase in sal, temperature increases by approximately 0.474 degrees.

--> Both coefficients are highly significant (p-value < 2e-16), suggesting strong evidence that sal is associated with temp.


### Residuals

- The median residual is close to 0 → good.

- Residuals are wide-spread from -19 to +30 degrees --> big variation


### Model Fit:

- Residual standard error: 6.683 → The typical error in predictions is about 6.68 units of temperature.

- R-squared: 0.2049 → Only ~20.5% of the variation in temp is explained by sal.

- F-statistic: 862,800 (huge) with p-value < 2.2e-16 → The model as a whole is statistically significant.


# Conclusion

- The relationship between salinity and temperature is statistically significant, and temperature increases as salinity increases.

- However, salinity only explains about 20.5% of the variation in temperature, so other factors are likely involved.

- The model is very precise (tiny standard errors), but not very explanatory — common in large datasets with complex systems